---
title: "NYPD Shooting Incidents Data Analysis"
author: "Engin Deniz Dogu"
date: "2024-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NYPD Shooting Data
#### Reading data...

```{r read_data, echo=TRUE}
nypd_shooting_data <- readr::read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

#### Data cleaning steps...

```{r data_cleaning, echo=TRUE}
# Date conversion
nypd_shooting_data$OCCUR_DATE <- as.Date(nypd_shooting_data$OCCUR_DATE, "%m/%d/%Y")

# Calculate total incidents in each BORO
total_incidents_in_boro <- nypd_shooting_data |> dplyr::group_by(BORO) |> 
  dplyr::summarize(cases=dplyr::n(),cases_per_thou = dplyr::n()/1000,
                   deaths_per_thou=sum(STATISTICAL_MURDER_FLAG==TRUE)/1000)

# Daily cases
cases_and_deaths_daily <- nypd_shooting_data |> dplyr::group_by(OCCUR_DATE) |>
  dplyr::summarize(cases=dplyr::n(),cases_per_thou = dplyr::n()/1000,
                   deaths_per_thou=sum(STATISTICAL_MURDER_FLAG==TRUE)/1000)

# Data Exploration (unique values)
boro_uniq <- unique(nypd_shooting_data["BORO"])
perp_race_uniq <- unique(nypd_shooting_data["PERP_RACE"])
perp_sex_uniq <- unique(nypd_shooting_data["PERP_SEX"])
perp_age_group_uniq <- unique(nypd_shooting_data["PERP_AGE_GROUP"])
vic_race_uniq <- unique(nypd_shooting_data["VIC_RACE"])
vic_sex_uniq <- unique(nypd_shooting_data["VIC_SEX"])
vic_age_group_uniq <- unique(nypd_shooting_data["VIC_AGE_GROUP"])
location_desc_uniq <- unique(nypd_shooting_data["LOCATION_DESC"])

# Discovery & Factors
boro_f <- c(unique(nypd_shooting_data$BORO))

perp_race_f <- subset(perp_race_uniq,!is.na(perp_race_uniq$PERP_RACE) 
                      & perp_race_uniq$PERP_RACE != "(null)")$PERP_RACE

perp_sex_f <- subset(perp_sex_uniq,!is.na(perp_sex_uniq$PERP_SEX) 
                      & perp_sex_uniq$PERP_SEX != "(null)")$PERP_SEX

perp_age_group_uniq_f <- subset(perp_age_group_uniq,!is.na(perp_age_group_uniq$PERP_AGE_GROUP) 
                      & perp_age_group_uniq$PERP_AGE_GROUP != "(null)"
                      & perp_age_group_uniq$PERP_AGE_GROUP != "1020"
                      & perp_age_group_uniq$PERP_AGE_GROUP != "940"
                      & perp_age_group_uniq$PERP_AGE_GROUP != "224"
                      & perp_age_group_uniq$PERP_AGE_GROUP != "1028"
                      )$PERP_AGE_GROUP

vic_race_uniq_f <- subset(vic_race_uniq,!is.na(vic_race_uniq$VIC_RACE) 
                      & vic_race_uniq$VIC_RACE != "(null)")$VIC_RACE

vic_sex_uniq_f <- subset(vic_sex_uniq,!is.na(vic_sex_uniq$VIC_SEX) 
                      & vic_sex_uniq$VIC_SEX != "(null)")$VIC_SEX

vic_age_group_uniq_f <- subset(vic_age_group_uniq,!is.na(vic_age_group_uniq$VIC_AGE_GROUP) 
                      & vic_age_group_uniq$VIC_AGE_GROUP != "(null)"
                      & vic_age_group_uniq$VIC_AGE_GROUP != "1022"
                      )$VIC_AGE_GROUP

location_desc_uniq_f <- subset(location_desc_uniq,!is.na(location_desc_uniq$LOCATION_DESC) 
                      & location_desc_uniq$LOCATION_DESC != "(null)"
                      )$LOCATION_DESC

# Remove unwanted columns
nypd_shooting_data <- subset(nypd_shooting_data,
                             select=-c(Lon_Lat,X_COORD_CD,Y_COORD_CD,JURISDICTION_CODE,OCCUR_TIME))

# Remove unwanted rows
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$PERP_RACE %in% perp_race_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$PERP_SEX %in% perp_sex_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$PERP_AGE_GROUP %in% perp_age_group_uniq_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$VIC_RACE %in% vic_race_uniq_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$VIC_SEX %in% vic_sex_uniq_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$VIC_AGE_GROUP %in% vic_age_group_uniq_f)
nypd_shooting_data <- subset(nypd_shooting_data,nypd_shooting_data$LOCATION_DESC %in% location_desc_uniq_f)

# Summary and missing data check
summary(nypd_shooting_data)
```

```{r analysis, echo=TRUE}
shooting_totals <- nypd_shooting_data |> dplyr::group_by(BORO, OCCUR_DATE) |>
  dplyr::summarize(cases=sum(INCIDENT_KEY), deaths=sum(STATISTICAL_MURDER_FLAG==TRUE))
  
shooting_totals <- subset(shooting_totals,deaths>0)
```

#### Total Cases by BORO

```{r histogram, echo=TRUE}
ggplot2::ggplot(nypd_shooting_data, ggplot2::aes(x=BORO,fill=BORO)) + ggplot2::geom_bar() + ggplot2::scale_fill_brewer(palette="Greens")
```

#### Comparison of Daily Deaths in MANHATTAN & BROOKLYN
```{r dot plot, echo=TRUE}
boro_selection <- "MANHATTAN"
selection <- subset(shooting_totals, BORO==boro_selection)
ggplot2::ggplot(selection, ggplot2::aes(x = OCCUR_DATE, y = deaths)) + ggplot2::geom_col()

boro_selection <- "BROOKLYN"
selection <- subset(shooting_totals, BORO==boro_selection)
ggplot2::ggplot(selection, ggplot2::aes(x = OCCUR_DATE, y = deaths)) + ggplot2::geom_col()
```

#### Number of Cases in Staten Island by Location Type
It seems that public housing, private housing and apartment buildings have the most shooting cases which might be interesting to look into.

```{r loc, echo=TRUE}
cases_by_location <- nypd_shooting_data |> dplyr::group_by(BORO,LOCATION_DESC) |> dplyr::summarize(n = dplyr::n())
  
boro_selection <- "STATEN ISLAND"
selection <- subset(cases_by_location, BORO==boro_selection)
ggplot2::ggplot(selection, ggplot2::aes(x = n, y = LOCATION_DESC)) + ggplot2::geom_col()
```

#### Modelling 1
Deaths per Thousand by BORO (predicted values in red).
Deaths per thousand is highly correlated with the number of cases.
```{r linear_regression_1, echo=TRUE}
model_1 <- lm(deaths_per_thou ~ cases_per_thou, data=total_incidents_in_boro)

summary(model_1)

total_incidents_in_boro_pred <- total_incidents_in_boro |> 
  dplyr::mutate(pred=predict(model_1))

total_incidents_in_boro_pred |> ggplot2::ggplot() +
  ggplot2::geom_point(ggplot2::aes(x=cases_per_thou, y=deaths_per_thou), color="blue") +
  ggplot2::geom_point(ggplot2::aes(x=cases_per_thou, y=pred),color="red")
```

#### Modelling 2
Deaths per Thousand by Date (predicted values in red).
It seems listing by date does not produce a good fit!
```{r linear_regression_2, echo=TRUE}
model_2 <- lm(deaths_per_thou ~ cases_per_thou, data=cases_and_deaths_daily)

summary(model_2)

cases_and_deaths_daily_pred <- cases_and_deaths_daily |> 
  dplyr::mutate(pred=predict(model_2))

cases_and_deaths_daily_pred |> ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x=cases_per_thou, y=deaths_per_thou), color="blue") +
  ggplot2::geom_point(ggplot2::aes(x=cases_per_thou, y=pred),color="red")
```

#### Sources of Bias and Final Thoughts
Potential sources of bias in NYPD shooting data:

 - Gender
 - Race
 - Age
 - Location
  
When I first looked at the I thought locations like bars/night clubs would have more shooting cases. However the seems to point at a different direction. It seems that most of the shootings are located in houses and apartments. My bias was that I considered houses to be safe. However, when I look at the analysis high numbers of shooting cases in houses and apartments actually make sense. Since these are private spaces (you wouldn't have cameras, police nearby etc.) it is possible to have a higher number of shooting cases related to these areas.

#### Session Summary

```{r session, echo=FALSE}
sessionInfo()
```